Rates of true polar wander in convecting planets

### 2015-01-26 tpw_rate.tex
Work on intro

\documentclass[extra,onecolumn]{gji}
%\documentclass[extra]{gji}
\usepackage{timet}
\usepackage{url}
\usepackage{amsmath}
\usepackage{graphicx}

@@

@@

The structure of thermal convection is primarily controlled by the Rayleigh number.  Once the Rayleigh number is sufficiently high ($\sim10^6$) 
the style of convection changes from steady/quasisteady to chaotic.  Accompanying this transition to chaos is the broadening of the spatial 
and temporal spectra \citep{mclaughlin1982transition}.

@@

The preceding results clarify the complex relationship between the Rayleigh number, the Froude number and the rates of TPW for a convecting planet.
As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principle axes of the convective moment of inertia. 
There is a constant competition between growth of the mismatch angle $\theta$ through Equation \ref{eq:grow_perturbation} and its relaxation through Equation \ref{simple_milankovitch}
We may plug in the estimate for $\Lambda_{ij}$ into equation \ref{simple_milankovitch} to find the strikingly simple expression
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{1}{\mathrm{Fr}} \sin{2 \theta}
\label{eq:simplest_milankovitch}
\end{equation}
Surprisingly, $\Gamma$ and $\mathrm{Ra}$ have completely dropped from the scaling. 
This is somewhat of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
However, we expect that more complicated estimates for $d(\mathrm{Ra})$ will still show a weak dependence on the Rayleigh number.

>>>>

%\documentclass[extra,onecolumn]{gji}
\documentclass[extra,mreferee]{gji}
\usepackage{timet}
\usepackage{url}
\usepackage{amsmath}
\usepackage{graphicx}

@@

A rotating, quasistatic body like a planetary mantle will tend to spin about the axis of its maximum moment of inertia.
Convection in a planetary mantle continuously redistributes mass, which can change the moment of ineratia tensor, necessitating a change in the spin axis of the planet to conserve angular momentum, a process known as true polar wander (TPW).

TPW was first considered in detail by \citet{darwin1887influence}, and the theory has been subsequently developed by many \citep[e.g.][]{munk1960rotation, goldreich1969some, ricard1993polar}. 
Despite this, the ability of internal mass anomalies to drive large-scale TPW remains controversial. 
Paleomagnetic data have been interpreted to requre up to $\sim 1^\circ$/Myr rates of TPW \citep{mitchell2011sutton}, but the ability of the mantle to respond at such rates has been questioned \citep{tsai2007theoretical}.

The primary uncertainties in assigning a maximum TPW rate to a convecting planet are the size of convective anomalies, which drive the rotational adjustment, and the viscosity structure of the mantle, which retards it. 
These two uncertainties are not unrelated: they are both expected to be functions of the geometric and material properties of the mantle.
As such, they do not vary independently. 
This suggests that an approach rooted in dimensional analysis and fluid dynamics can clarify the rates and magnitudes of TPW.

Most previous studies coupling mantle convection models to polar wander calculations have done so with prescribed density perturbations \citep[e.g.][]{greff2004upwelling}, or prescribed moment if inertia variations \citep[e.g.]{tsai2007theoretical, creveling2012mechanisms}. 
\citet{richards1999polar} coupled full thermal convection models to a polar wander model, but did not address in detail the scaling relationships between the two.

Herein we perform a scaling analysis for a minimal system of a rotating, convecting mantle.

@@

The structure of thermal convection is primarily controlled by the Rayleigh number.  Once the Rayleigh number is sufficiently high ($\sim10^6$) 
the style of convection changes from steady/quasisteady to chaotic.  Accompanying this transition to chaos is a broadening of the spatial 
and temporal spectra \citep{mclaughlin1982transition}.

@@

The preceding results clarify the complex relationship between the Rayleigh number, the Froude number and the rates of TPW for a convecting planet.
As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principle axes of the convective moment of inertia. 
There is a constant competition between growth of the mismatch angle $\theta$ through Equation \ref{eq:grow_perturbation} and its relaxation through Equation \ref{simple_milankovitch}.
We may plug in the estimate for $\Lambda_{ij}$ into equation \ref{simple_milankovitch} to find the strikingly simple expression
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{1}{\mathrm{Fr}} \sin{2 \theta}
\label{eq:simplest_milankovitch}
\end{equation}
Surprisingly, $\Gamma$ and $\mathrm{Ra}$ have completely dropped from the scaling. 
This is somewhat of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
However, we expect that more complicated estimates for $d(\mathrm{Ra})$ will still show a weak dependence on the Rayleigh number.
